Static contract checking for Haskell

### 2011-12-22 paper/paper.tex
Paper: high level description of 'min' and many smaller edits.
Still need to motivate the particular axioms, and explain the proof/use
distinction in contract translation.

@@

We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
  CF(fst~(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
can't think of a good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons~y~ys \neq \unr
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\phi_\text{lazy}$) say that constructors are ``lazy'' or ``total''.}

Finally, we define projections for $Cons$. \annote[nc]{It is not
  strictly necessary, but it will be handy:}{??? Why is it handy? It
  doesn't seem to play any role in formal part of this paper (besides
  being specified in axioms $\phi_\text{injective}$). In the current
  implementation it's used to avoid introducing existentially
  quantified variables in the $\bigvee_i$ part of $\dtrans{}$ for
  functions defined by case expressions. Equation \eqref{unrnotnull}
  below uses it that way. ??? Does Equinox care?}
\begin{equation*}
\forall xs,y,ys.~\proj{1}{Cons}~Cons~y~ys = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~\proj{2}{Cons}~Cons~y~ys = xs \implies xs = ys 
\end{equation*}\note[nc]{??? Why not the simpler \[\forall y,ys.~\proj{2}{Cons}~Cons~y~ys = ys\]? Same question in Figure \ref{dtrans}.  (Equinox does have a built in notion of equality, e.g. running equinox on
\center{\texttt{fof(\_,theorem,? [X,Y] : (X = Y \& f(X) != f(Y))).}}\\
 results in ``Unsatisfiable'', so support for equality reasoning is not the reason.)}
%% \[
%% \text{\texttt{\$ echo "fof(\_,theorem,? [X,Y] : (X = Y & f(X) != f(Y)))."}} \\
%%                  \text{\texttt{| equinox /dev/stdin}}}}
%% \]

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\footnote{
%
I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.
%
}
The expression language in our logic is the expression language of 
\hprime (Figure~\ref{fig:hprime-syntax}).  Note that curried application in \hprime
corresponds to a binary function symbol in FOL, which we denote by juxtaposition.

@@

\section{Translations}
For an overview of the different translations we define, see figure~\ref{trans}

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \phi~(\mbox{formula}) \\
    \ttrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formula}) \\
    \ctrans{f \in c} &\to& \phi~(\mbox{formula}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{trans}
\end{figure}

@@

\subsection{$\dtrans{}$ -- Definitions}
We give in figure~\ref{dtrans} the translation of function
definitions. The first line says that when applied to an argument that
matches a pattern of the case expression, we should equate the
function call to the corresponding expression. The second line states
that if the pattern-matching failed or if we pattern-matched on $\bad$
then the result should be $\unr$.\note[nc]{This text does not agree
  with the figure.  In particular, the axiom says nothing about the
  result of scrutinizing $\bad$.  I think the correct behavior is that
  scrutinizing $\bad$ results in $\bad$, \emph{not} $\unr$.}

\newcommand{\fullfxs}{\full{f}(\vec{x})}
\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f~\vec{x} = e} &=& \forall \vec{x}.\min{\fullfxs} 
                                              \to \fullfxs = e \\
   \raisebox{ 0.5 cm }{$\dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}$} &\raisebox{0.5cm}{=}&  \hspace{-0.18cm} \begin{array}{rclcl}
     \forall \vec{x}. \min{\fullfxs} &\to&
     \multicolumn{3}{c}{  \min{e} \land (e = \texttt{BAD} \lor
       \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i})\right) \lor \fullfxs = \texttt{UNR})} \\
     &\land& \forall \vec{x_1}.~e = \full{K}_1(\vec{x_1}) &\to& \fullfxs = e_1 \\
     &\land& & \cdots & \\
     &\land& \forall \vec{x_n}.~e = \full{K}_n(\vec{x_n}) &\to& \fullfxs = e_n \\
   \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Defintions translation}
 \label{dtrans}
 \note[nc]{??? Why do we introduce the extra quantified variable $a$,
   representing the result (``answer''?), instead of just using the result directly
   (which is logically equivalent)? Same question for the next
   Figure.  Does this make a difference to Equinox?}
\end{figure*}

\subsection{$\ttrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in Figure~\ref{ktrans}
$$\ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$

\begin{itemize}
\item $(\phi_\text{injective})$ Constructors are injective.  That is,
  for each term constructor $K$, if $K e_1 \cdots e_n = K e'_1 \cdots
  e'_n$, then $e_i = e'_i$ for all $i$.  We axiomitize this via
  projections $\proj{i}{K}$: for each $i$, $\proj{i}{K}(K e_1 \cdots
  e_n) = e_i$.  NB: unlike for user defined functions, the $\min{}$
  constraint on unfolding projections is on the \emph{argument}, and
  not the \emph{function application} itself.  I.e., instead of
  $\min{\proj{i}{K}(K e_1 \cdots e_n)}$ we have $\min{K e_1 \cdots
    e_n}$.  We guard with the latter, not the former, because the
  primary purpose of projectors is to axiomitize the injectivity of
  constructors.  E.g., given $S~x_1 = S~x_2$, and $\min{S~x_1}$, we
  want to conclude that $x_1 = x_2$.  We won't, in general, have
  $\min{\proj{1}{S}(S~x_1)}$ here, so such a min-constraint would
  block us.

  We also use projections in some versions of the case-expression
  translation, to Skolemize, but this is just a happy coincidence.  In
  particular, user programs will never contain projections.  If a user
  wants to project then they need to write there own projection
  function using pattern matching.
\item $(\phi_\text{disjoint})$ Distinct term constructors have
  disjoint ranges.  Note that we don't axiomitize disjointness of term
  constructors from distinct \emph{data types}.
\item $(\phi_\text{cf})$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  If $K$ has arity $n$ then we
    assert $\ctrans{K \in \cfc^n \to \cfc}^-$.  I.e., we simply treat
    K as a strict \cfc function.

@@

  Note that $\phi_\text{cf}$ is not true for arbitrary functions: a function is not required
  to use all of its arguments. For example, \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.

@@

\item $(\phi_\text{lazy})$ No full application of a constructor
  $K_i$ is unreachable or crashing.
\end{itemize}

\begin{figure*}
  $$ \ruleform{\ttrans{data~def} = \Phi} $$
$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phi_\text{injective} \land \phi_\text{disjoint} \land \phi_\text{cf} \land \phi_\text{lazy}$$
 \hspace{5.6cm}where
  \begin{center}
    \newcommand{\fullkxs}[1]{\full{K}_#1~\vec{x}}
    \begin{align*}
      \phi_\text{injective} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
      \phi_\text{disjoint} &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
      \phi_\text{cf} &= 
        \bigwedge_{1 \leq i \leq n}~\left( \ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
          \land \forall \vec{x}.~\min{\fullkxs i}
                  \land \cf{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} \cf{x_j} \right) \\
      \phi_\text{lazy} &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
    \end{align*}
  \end{center}
  \caption{$\ttrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
in Figure~\ref{dtrans}.  Definition translations are axioms, 
so the the initial $\min{\full{f}(\vec{x})}$
is an obligation: we aren't allowed to 
``unfold the definition'' of the function unless we prove
that we're interested in a particular application.  
Once that $\min$ obligation is satisfied, interest ($\min{e}$) in the
scrutinee of the case match may be concluded, because the function
must inspect the scrutinee to compute the result.

@@

\begin{figure*}
\begin{verbatim}
[[e in {x|p}]]      := e /= unr -> p[e/x] in {true, unr}
[[e in cf]]         := cf(e)
[[e in x:c1 -> c2]] := forall x. [[x in c1]] -> [[e x in c2]]
\end{verbatim}
\caption{$\ctrans{}$ -- Naive (no $\min$) contract translation.}
\label{fig:naiveCTrans}
\end{figure*}

\begin{figure*}
\begin{verbatim}
-- Need 'min(e)' and 'min(p[e/x])' here to unfold defs.
[| e ::: {x|p} |]+ = min(e) /\ min(p[e/x]) ->
                     (e /= unr -> p[e/x] not-in {false, bad})
-- Need 'min(e)' here to guard 'min(p[e/x])'.
-- Need 'min(p[e/x])' here to unfold defs.
[| e ::: {x|p} |]- = min(e) -> min(p[e/x]) /\
                     (e /= unr -> p[e/x]     in {true, unr})

-- Need 'min(e)' here in the (+)-case to unfold defs.
-- The 'min(e)' in the (-)-case doesn't seem strictly
-- necessary, but it gives a big performance improvement for
-- 'yes/add-symmetric.hs', from 48s to 30s for a full check.
[[e in cf]]v := min(e) ->  cf(e)

 -- No auxiliary mins.
[[e in x:c1 -> c2]]v
  := forall x. [[x in c1]](dual v) -> [[e x in c2]]v
\end{verbatim}
\caption{$\ctrans{}$ -- New contract translation. Design choices, in the form
of optional parts, are enclosed in single square brackets.}
\label{fig:newCTrans}
\end{figure*}

@@

Going back to the current translation, we consider some of its
metatheoretic properties.
\protect \begin{enumerate}
\item A standard
  property of bidirectional systems: that you can check/verify what you assume/use.
  In other words, forall $e$ and $c$,
  \protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
  \]
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' translation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like: given some validity or consistency 
  assumptions on $\Gamma$ and $e \in c$,
  \protect \[
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.
\end{enumerate}

@@

  \item \add[nc]{state that $\unr$ is crash-free: $\cf{\unr}$.}\note[nc]{
This is already present in the implementation, and assumed in the
consistency proof.}

  \item give formulae for the boolean datatype, equivalent to
translating
\begin{verbatim}
data Bool = True 0 | False 0
\end{verbatim}

@@

\begin{figure}\label{fig:prelude}
  \begin{itemize}
    \item $\lnot \cf{\bad}$
    \item $\cf{\unr}$
    \item $\dtrans{data~Bool = \true~0 | \false~0}$
    \item \designChoice\\ $\forall f,x. \min{f ~x} \to \min{f}$
    \item \designChoice\\ $\forall x. \min{\bad ~x} \to \bad ~x = \bad$ 
    \item \designChoice\\ $\forall x. \min{\unr ~x} \to \unr ~x = \unr$
    \item \designChoice\\ $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$
    \item 
  \end{itemize}
  \caption{Prelude -- Axioms common to all translations.}
\end{figure}

>>>>

\newcommand{\phiLazy}{\phi_\text{lazy}}
\newcommand{\phiCF}{\phi_\text{CF}}
\newcommand{\phiInjective}{\phi_\text{injective}}
\newcommand{\phiDisjoint}{\phi_\text{disjoint}}

@@

We also say some stuff about \annote[nc]{unreachibility}{??? e.g. \[
  CF(fst~(x,BAD)) \iff CF(x) \] as in section \ref{sec:cf}?} but I
can't think of a good way to explain it right now.
\begin{equation*}
\forall y,ys.~Cons~y~ys \neq \unr
\end{equation*}
\begin{equation*}
Nil \neq \unr 
\end{equation*}\note[nc]{\unr means ``infinite loop'', approximately (it's also the result of pattern matches on non-loop expressions with the wrong constructor head), and so these rules ($\phiLazy$) say that constructors are ``lazy'' or ``total''.}

Finally, we define projections for $Cons$. \annote[nc]{It is not
  strictly necessary, but it will be handy:}{??? Why is it handy? It
  doesn't seem to play any role in formal part of this paper (besides
  being specified in axioms $\phiInjective$). In the current
  implementation it's used to avoid introducing existentially
  quantified variables in the $\bigvee_i$ part of $\dtrans{}$ for
  functions defined by case expressions. Equation \eqref{unrnotnull}
  below uses it that way. ??? Does Equinox care?}
\begin{equation*}
\forall xs,y,ys.~\proj{1}{Cons}~Cons~y~ys = xs \implies xs = y 
\end{equation*}
\begin{equation*}
\forall xs,y,ys.~\proj{2}{Cons}~Cons~y~ys = xs \implies xs = ys 
\end{equation*}\note[nc]{??? Why not the simpler \[\forall y,ys.~\proj{2}{Cons}~Cons~y~ys = ys\]? Same question in Figure \ref{fig:dtrans}.  (Equinox does have a built in notion of equality, e.g. running equinox on
\center{\texttt{fof(\_,theorem,? [X,Y] : (X = Y \& f(X) != f(Y))).}}\\
 results in ``Unsatisfiable'', so support for equality reasoning is not the reason.)}
%% \[
%% \text{\texttt{\$ echo "fof(\_,theorem,? [X,Y] : (X = Y & f(X) != f(Y)))."}} \\
%%                  \text{\texttt{| equinox /dev/stdin}}}}
%% \]

@@

\subsection{First-order logic with equality}
We use first-order logic with equality, defined in figure \ref{fol-stx}.\footnote{
%
I guess ``with equality'' means that the proof theory has an axiom schema for substition, i.e. forall $\phi$, $\forall x,y.x=y \to (\phi \tot \phi[y/x])$, axioms making $=$ an equality relation, and the requirement that all models interpret $=$ as true equality on the carrier.
%
}
The expression language in our logic is the expression language of 
\hprime (Figure~\ref{fig:hprime-syntax}).\footnote{
  %
  Note that this does \emph{not} include \texttt{case} expressions.
  %
}  Note that curried application in \hprime
corresponds to a binary function symbol in FOL, which we denote by juxtaposition.

@@

\section{Translations}
For an overview of the different translations we define, see
Figure~\ref{fig:all-trans}.

\begin{figure}
 \begin{center}
  \[  \begin{array}{rcl}
    \dtrans{def} &\to& \phi~(\mbox{formula}) \\
    \ttrans{\mbox{data } T = \dots} &\to& \phi~(\mbox{formula}) \\
    \ctrans{f \in c} &\to& \phi~(\mbox{formula}) \\
  \end{array} \]
  \end{center}
  \caption{Translations}
  \label{fig:all-trans}
\end{figure}

\subsection{The $\min$ predicate}

We introduce a predicate $\min$, short for ``am interested'', and use
it both to guide proof search and guarantee (?) the existence of
finite models.  Dimitrios says that such use of a predicate, at least
the guiding of proof search part, is related to the concept of
``triggers''.  We are particularly interested in finite
\emph{counter}-models, which exist when the contracts aren't provable
(i.e. when our refutation theory is counter satisfiable).

The $\min$ predicate guides proof search by restricting the use of
axioms. Axioms are guarded by $\min$ preconditions, and in some cases
introduce new $\min$s.  The rough idea is that $\min$ corresponds to
evaluation: you have $\min(e)$ when you force the value of $e$ in our
lazy language.  For example, the axioms for function definitions in
Figure~\ref{fig:dtrans} require interest $\min{f x}$ to unfold the
definition of $f$, but then introduce $\min{e}$ if $f$ is defined by
\texttt{case}-matching $e$, because \texttt{case}-matching forces the
scrutinee.

The $\min$ predicate guarantees finite models by preventing us from
proving all models are infinite (recall the completeness theorem for
FOL).  Without $\min$, the axiomitization of any infinite datatype $T$
would allow us to prove $T$, and hence the domain of our models, was
infinite.  Namely, we axiomitize the injectivity and disjointness of
$T$'s term constructors, and without $\min$ these axioms prove that
inductive term constructors are injective, but not surjective, and
hence the domain is infinite.  However, we guard these axioms, namely
$\phiInjective$ and $\phiDisjoint$ in Figure~\ref{fig:dtrans}, with
$\min$ predicates, to restrict their use to interesting constructor
applications.

@@

\subsection{$\dtrans{}$ -- Definitions}
We give in figure~\ref{fig:dtrans} the translation of function
definitions. The first line says that when applied to an argument that
matches a pattern of the case expression, we should equate the
function call to the corresponding expression. The second line states
that if the pattern-matching failed or if we pattern-matched on $\bad$
then the result should be $\unr$.\note[nc]{This text does not agree
  with the figure.  In particular, the axiom says nothing about the
  result of scrutinizing $\bad$.  I think the correct behavior is that
  scrutinizing $\bad$ results in $\bad$, \emph{not} $\unr$.}

\newcommand{\fullfxs}{\full{f}(\vec{x})}
\begin{figure*}
\[
\begin{array}{rcl}
  \multicolumn{3}{c}{\ruleform{\dtrans{def} = \phi}} \\[2.5mm]
  \dtrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} &=& \ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} \\
  \dtrans{f \in c} &=& \ctrans{f \in c}\\
  \dtrans{f~\vec{x} = e} &=& \forall \vec{x}.\min{\fullfxs} 
                                              \to \fullfxs = e \\
   \dtrans{f~\vec{x} = \mbox{case } e \mbox{ of } [K_i~\vec{x_i} = e_i]}
    &=&
    \begin{array}{ll}
      \forall \vec{x}. \min{\fullfxs} 
      &\to \min{e} \\
      &\land \left(
        \bigwedge_i \left(
                      \forall \vec{x_i}.~e = \full{K}_i(\vec{x_i}) \to \fullfxs = e_i
                    \right)
             \right) \\
      &\land \left( e = \bad \to \fullfxs = \bad \right) \\
      &\land \left( e = \bad \lor 
                    \left(\bigvee_i \exists \vec{x_i}.~e = \full{K}_i(\vec{x_i}) \right) 
                    \lor \fullfxs = \unr
             \right) \\
    \end{array} 
\end{array}
\]
\caption{$\dtrans{}$ -- Definition translation}
 \label{fig:dtrans}
\end{figure*}

\subsection{$\ttrans{}$ -- Datatypes}
We break down the translation for datatypes in four parts, described in Figure~\ref{ktrans}
$$\ttrans{\mbox{data } T = K_1~a_1, \dots, K_n~a_n} = \phiInjective \land \phiDisjoint \land \phiCF \land \phiLazy$$

\begin{itemize}
\item $(\phiInjective)$ Constructors are injective.  That is,
  for each term constructor $K$, if $K e_1 \cdots e_n = K e'_1 \cdots
  e'_n$, then $e_i = e'_i$ for all $i$.  We axiomitize this via
  projections $\proj{i}{K}$: for each $i$, $\proj{i}{K}(K e_1 \cdots
  e_n) = e_i$.  NB: unlike for user defined functions, the $\min{}$
  constraint on unfolding projections is on the \emph{argument}, and
  not the \emph{function application} itself.  I.e., instead of
  $\min{\proj{i}{K}(K e_1 \cdots e_n)}$ we have $\min{K e_1 \cdots
    e_n}$.  We guard with the latter, not the former, because the
  primary purpose of projectors is to axiomitize the injectivity of
  constructors.  E.g., given $S~x_1 = S~x_2$, and $\min{S~x_1}$, we
  want to conclude that $x_1 = x_2$.  We won't, in general, have
  $\min{\proj{1}{S}(S~x_1)}$ here, so such a min-constraint would
  block us.

  We also use projections in some versions of the case-expression
  translation, to Skolemize, but this is just a happy coincidence.  In
  particular, user programs will never contain projections.  If a user
  wants to project then they need to write there own projection
  function using pattern matching.
\item $(\phiDisjoint)$ Distinct term constructors have
  disjoint ranges.  Note that we don't axiomitize disjointness of term
  constructors from distinct \emph{data types}.
\item $(\phiCF)$ Then, we have to give crash-freeness conditions for each $K_i$:
  Notice that we have a equivalence.
  \begin{itemize}
  \item $\gets$: if we pack crash-free values in a data constructor,
    the resulting value is crash-free.  If $K$ has arity $n$ then we
    assert $\ctrans{K \in \cfc^n \to \cfc}^-$.  I.e., we simply treat
    K as a strict \cfc function.

@@

  Note that $\phiCF$ is not true for arbitrary functions: a function is not required
  to use all of its arguments. For example, \texttt{fst} is crash-free if and only
  if the first argument of the pair is crash-free. The second argument
  is irrelevant.

@@

\item $(\phiLazy)$ No full application of a constructor
  $K_i$ is unreachable or crashing.
\end{itemize}

\begin{figure*}
  $$ \ruleform{\ttrans{data~def} = \Phi} $$
$$\ttrans{\mbox{data } T = K_1 ~a_1, \dots, K_n ~a_n} = \phiInjective \land \phiDisjoint \land \phiCF \land \phiLazy$$
 \hspace{5.6cm}where
  \begin{center}
    \newcommand{\fullkxs}[1]{\full{K}_#1~\vec{x}}
    \begin{align*}
      \phiInjective &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~\min{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} x_j = \proj{j}{K_i}(\fullkxs i) \right) \\
      \phiDisjoint &= \!\!\!\bigwedge_{1 \leq i < j \leq n}\!\!\!~\left(\forall \vec{x},\vec{y}.~ (\min{\fullkxs i} \lor \min{\fullkxs j}) \to \fullkxs {i} \neq \fullkxs {j} \right)\\
      \phiCF &= 
        \bigwedge_{1 \leq i \leq n}~\left( \ctrans{K_i \in \cfc^{a_i} \to \cfc}^- 
          \land \forall \vec{x}.~\min{\fullkxs i}
                  \land \cf{\fullkxs i} \to \bigwedge_{1 \leq j \leq a_i} \cf{x_j} \right) \\
      \phiLazy &= \bigwedge_{1 \leq i \leq n}~\left( \forall \vec{x}.~(\min{\fullkxs i} \to \fullkxs {i} \neq \unr \land \fullkxs {i} \neq \bad \right)
    \end{align*}
  \end{center}
  \caption{$\ttrans{}$ -- Data type translation}
  \label{ktrans}
\end{figure*}

@@

As another example, consider the $\dtrans{}$ for functions defined by pattern matching,
in Figure~\ref{fig:dtrans}.  Definition translations are axioms, 
so the the initial $\min{\full{f}(\vec{x})}$
is an obligation: we aren't allowed to 
``unfold the definition'' of the function unless we prove
that we're interested in a particular application.  
Once that $\min$ obligation is satisfied, interest ($\min{e}$) in the
scrutinee of the case match may be concluded, because the function
must inspect the scrutinee to compute the result.

@@

\begin{figure*}
\begin{verbatim}
[|e in {x|p}|]      := e /= unr -> p[e/x] in {true, unr}
[|e in cf|]         := cf(e)
[|e in x:c1 -> c2|] := forall x. [|x in c1|] -> [|e x in c2|]
\end{verbatim}
\caption{$\ctrans{}$ -- Naive (no $\min$) contract translation.}
\label{fig:naiveCTrans}
\end{figure*}

\begin{figure*}
\begin{verbatim}
-- Need 'min(e)' and 'min(p[e/x])' here to unfold defs.
[| e in {x|p} |]+ = min(e) /\ min(p[e/x]) ->
                      (e /= unr -> p[e/x] not-in {false, bad})
-- Need 'min(e)' here to guard 'min(p[e/x])'.
-- Need 'min(p[e/x])' here to unfold defs.
[| e in {x|p} |]- = min(e) -> min(p[e/x]) /\
                      (e /= unr -> p[e/x]     in {true, unr})

-- Need 'min(e)' here in the (+)-case to unfold defs.
-- The 'min(e)' in the (-)-case doesn't seem strictly
-- necessary, but it gives a big performance improvement for
-- 'yes/add-symmetric.hs', from 48s to 30s for a full check.
[|e in cf|]v := min(e) ->  cf(e)

 -- No auxiliary mins.
[|e in x:c1 -> c2|]v
  := forall x. [|x in c1|](dual v) -> [|e x in c2|]v
\end{verbatim}
\caption{$\ctrans{}$ -- New contract translation. Design choices, in the form
of optional parts, are enclosed in single square brackets.}
\label{fig:newCTrans}
\end{figure*}

@@

Metatheoretic properties of the translation
\begin{enumerate}
\item A standard
  property of bidirectional systems: that you can check/verify what you assume/use.
  In other words, forall $e$ and $c$,
  \protect \[
  \ctrans{e \in c}^- \text{ implies } \ctrans{e \in c}^+.
  \]
\item The ``fancy'' $\min{}$-ified translation should 
  degenerate to the ``naive'' translation when we assume $\forall x. \min{x}$.
  This is trivially true if we define the ``naive'' translatino as the ``min'' translation
  plus that axiom.  In symbols, we have
  \protect \[
  \Gamma^\text{min} \vdash \phi \implies \Gamma^\text{naive} \vdash \phi.
  \]  I.e., ``\min{} is sound for `naive'''.
\item We'd like to have a result  like: given some validity or consistency 
  assumptions on $\Gamma$ and $e \in c$,
  \[
    \Gamma^\text{naive} \vdash \ctrans{e \in c}^\text{naive} \implies \Gamma^\text{min} \vdash \ctrans{e \in c}^\text{min}.
  \]  I.e., ``\min{} is complete for `naive', under some assumptions about 
  what we're proving''.  However, Dimitrios points out that we shouldn't expect this
  completeness theorem without some work, since we're purposefully 
  restricting the proof search to follow
  evaulation strategy, and it's far from obvious that all proofs can proceed this way.
\end{enumerate}

@@

  \item state that $\unr$ is crash-free: $\cf{\unr}$.

@@

\begin{figure}\label{fig:prelude}
  The implementation enables
  \begin{itemize}
    \item $\lnot \cf{\bad}$
    \item $\cf{\unr}$
    \item \designChoice\\ $\forall f,x. \min{f ~x} \to \min{f}$
  \end{itemize}
  and disables
  \begin{itemize}
    \item \designChoice\\ $\forall x. \min{\bad ~x} \to \bad ~x = \bad$ 
    \item \designChoice\\ $\forall x. \min{\unr ~x} \to \unr ~x = \unr$
    \item \designChoice\\ $\forall f. \cfc(f) \to \ctrans{f \in \cfc \to \cfc}^-$
  \end{itemize}
  \caption{Prelude -- Axioms common to all translations.}
\end{figure}
